---
title: "Discussion 3"
author: "Sam Reeves"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(faraway)

data(teengamb, package = 'faraway')
tibble(teengamb)
```

# 3.3

> Using the teengamb data, fit a model with gamble as  response and the other variables as predictors.

```{r}
lmod <- lm(gamble ~ ., teengamb)
```

> (a) Which variables are statistically significant at the 5% level?

```{r}
summary(lmod)
```
Only income and sex!

> (b) What interpretation should be given to the coefficient for sex?

It has significant predictive value, however income will be a bigger deciding factor alone, and it will affect the predictive power of sex.

> (c) Fit a model with just income as a predictor and use an F-test to compare it to the full model.

```{r}
lmod2 <- lm(gamble ~ income, teengamb)
summary(lmod2)
```
```{r}
curve(df(x, df1=4, df2=42), from=0, to=4, xlab = '', ylab = '', col = 'blue', main = 'F Distributions - lmod2 in black', axes=FALSE)
par(new=TRUE)
curve(df(x, df1=1, df2=45), from=0, to=4, xlab = '', ylab = '', axes = FALSE)
```
```{r}
nullmod <- lm(gamble ~ 1, teengamb)
anova(nullmod, lmod)
anova(nullmod, lmod2)
```
The model with only income seems about a third as effective.
